Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2014-11-26 manuscript.tex
Joined text for the theory part with equations

where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 0.00000000006673\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
$\kappa = {r^\prime}^2 \cos \phi^\prime$,
and

@@

We will follow \citet{Asgharzadeh2007} and perform the numerical integration
using the Gauss-Legendre Quadrature (GLQ).

The GLQ consists of approximating the integral by a weighted sum of the
integration kernel \citep{Hildebrand1987},
\begin{equation}
    \int\limits_a^b f(x) dx \approx
    \frac{b-a}{2}\sum\limits_{i=1}^N W_i f(x_i),
    \label{eq:glq1d}
\end{equation}

$N$ is the order of the quadrature, i.e. the number of points used in the GLQ.

The points $x_i$ where the kernel is evaluated are called the quadrature nodes.

The points are roots of the $N^{th}$ order Legendre polynomial $P_N(x)$.

The roots are $x = \pm 0.577350269$ for a second order polynomial ($P_2(x)$).

A root finder algorithm can be used to determine the roots for larger order
polynomials.

The second order roots, as well as the ones determined by a root finder
algorithm, will be within the range $[-1, 1]$.

Before being used for GLQ integration,
the roots must be scaled to the integration limits $[a, b]$ using

@@

\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P^\prime_N(x_i))^2},
    \label{eq:glq_weights}
\end{equation}

The values of the $P_N(x)$ and its first derivative $P^\prime_N(x)$ can be
calculated with recursive relations.

The Gauss-Legendre Quadrature for volume integrals, like equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
becomes \citep{Asgharzadeh2007}

@@

where

@@

Applying equation~\ref{eq:glq3d} to equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
we see that $f(r_i, \phi_j, \lambda_k)$ is the effect of a point
mass located on the quadrature nodes.

It can be said that the GLQ integration approximates the volume integrals  by a
weighted sum of point mass effects.

The accuracy of the integration
depends on the number of point masses used in the summation.

\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.

Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.

The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.

The maps were calculated on a regular grid
with $100\times100$ points.

The GLQ order used was the same
in the radial, longitudinal, and latitudinal dimensions.

Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two (two point masses in the longitude, latitude, and vertical
directions, respectively).

Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.

Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.

\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.

Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.

One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.

Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30.

The computed $g_{xy}$ component more closely resembles the results from
\citet{Asgharzadeh2007}.

>>>>

\noindent
where $\alpha, \beta \in \{x, y, z\}$,
$\rho$ is the density,
$G = 0.00000000006673\ m^3kg^{-1}s^{-1}$ is the gravitational constant,
$\kappa = {r^\prime}^2 \cos \phi^\prime$,
and

@@

We will follow \citet{Asgharzadeh2007} and perform the numerical integration
using the Gauss-Legendre Quadrature (GLQ).
The GLQ consists of approximating the integral by a weighted sum of the
integration kernel \citep{Hildebrand1987},
\begin{equation}
    \int\limits_a^b f(x) dx \approx
    \frac{b-a}{2}\sum\limits_{i=1}^N W_i f(x_i),
    \label{eq:glq1d}
\end{equation}

\noindent
in which $N$ is the order of the quadrature,
i.e. the number of points used in the GLQ.
The points $x_i$ are called the quadrature nodes.
They are the roots of the $N^{th}$ order Legendre polynomial $P_N(x)$.
For a second order polynomial ($P_2(x)$),
the roots are $x = \pm 0.577350269$.
Roots for larger order polynomials
can be determined by a root finder algorithm.
Roots of Legendre polynomials
will be within the range $[-1, 1]$.
Before being used for GLQ integration,
the roots must be scaled to the integration limits $[a, b]$ using

@@

\begin{equation}
    W_i = \frac{2}{(1 - x_i^2)(P^\prime_N(x_i))^2}.
    \label{eq:glq_weights}
\end{equation}

\noindent
The values of $P_N(x)$ and its first derivative $P^\prime_N(x)$
can be calculated with recursive relations.

The Gauss-Legendre Quadrature for three-dimensional volume integrals,
like equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
becomes \citep{Asgharzadeh2007}

@@

\noindent
where

@@

Applying equation~\ref{eq:glq3d} to equations
\ref{eq:tesspot}, \ref{eq:tessgrav}, and \ref{eq:tesstensor},
we see that $f(r_i, \phi_j, \lambda_k)$ is the effect of a point
mass located on the quadrature nodes.
Thus, it can be said that the GLQ integration
approximates the volume integrals  by a
weighted sum of point mass effects.

The accuracy of the integration
depends on the number of point masses used in the summation.
\citet{Ku1977} showed that it also depends on the ratio between
the distance to the computation point and the distance between adjacent nodes.
Figure~\ref{fig:glqerrorsample}
illustrates this effect on the $g_{xy}$ gravity gradient component.
The $g_{xy}$ component was produced by a
$7^\circ \times 7^\circ \times 20\ km$ tesseroid
with $2.67\ g.cm^{-3}$ density
and top at $z=0\ km$.
The maps were calculated on a regular grid
with $100\times100$ points.
Figure~\ref{fig:glqerrorsample}a shows the $g_{xy}$ component
calculated at 400 km height using
GLQ with order two (two point masses in the longitude, latitude, and vertical
directions, respectively).
Figure~\ref{fig:glqerrorsample}b shows $g_{xy}$ computed with order two
GLQ as well but at 150 km height.
Notice that the computed effect is concentrated around each point mass
of the GLQ (black dots) and does not resemble the effect of a tesseroid.
\citet{Ku1977} determined an empirical criterion that the distance between
point masses (quadrature nodes) should be smaller than the minimum distance to
the computation point.
Thus, if a computation point is too close to the tesseroid one would have to
decrease the distance between the point masses in order to obtain an accurate
result.
One way to accomplish this would be increase the order of the quadrature
$N$ in all three directions.
Figure~\ref{fig:glqerrorsample}c shows the $g_{xy}$ component calculated at
150km height but with a GLQ order of 30.
The computed $g_{xy}$ component more closely resembles the results from
\citet{Asgharzadeh2007}.
